---
title: "DEseq2"
author: "David Requena and Jamie Saltsman"
date: "September 12, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Differential expression analysis using DEseq2

## 1. SET UP

### Installing and/or calling the libraries

This first block is just to install the dependencies. If already installed, just left this commented.

```{r Install libraries, eval=FALSE, echo=FALSE}
# # Import necessary libraries
# install.packages('ggplot2')
# install.packages('gplots')
# install.packages('RColorBrewer')
# install.packages('matrixStats')
# 
# source("https://bioconductor.org/biocLite.R")
# biocLite("XML") #In linux, first do: sudo apt-get install libxml2-dev
# biocLite("DESeq2")
# biocLite("biomaRt")
# biocLite("pheatmap")
```
This block is to call all the required libraries:

```{r, echo=FALSE}
library("XML")
library("DESeq2")
library("biomaRt")
library("gplots")
library("RColorBrewer")
library("ggplot2")
library("pheatmap")
library("matrixStats")
```

### Input sample metadata

In this block, you just need to change the working directory:

```{r "setup"}
require("knitr")
knitr::opts_knit$set(root.dir = "../RNAseq_data") # <<<------------- *** WORKING DIRECTORY ***

# The samples' data and the counts' table must be in your working directory
sample_data <- "samples_data.tsv"
counts <- "counts.tsv"
```

This block is to read the samples' data.
There is also an option to work with subsets.

```{r, echo=FALSE}
#########################
# Input sample metadata #
#########################

sampledata <- data.frame(read.csv(sample_data, sep = "\t", header = TRUE))

# The next two lines are only necessary if you're planning to work with a subset of the data:
#sampledata <- sampledata[sampledata$library_prep!=0,]
#sampledata <- sampledata[sampledata$have_npm=="Y",]
sampledata <- sampledata[sampledata$normal_pm=="Y",]
sampledata$library_prep <- factor(sampledata$library_prep)

# To sort the table by sample name:
sampledata <- sampledata[order(sampledata[,2]),]
sampledata$patient <- as.character(sampledata$patient)
rownames(sampledata) <- sampledata$id
```

This block is to read the tables with the counts.
There is also an option to filter out some genes (blacklist).
```{r, echo=FALSE}
######################
# Input counts table #
######################

countstable <- data.frame(read.csv(counts, sep = "\t", header = FALSE), stringsAsFactors = FALSE)

# Filter out some genes:
# black_list = c("ENSG00000142173", "ENSG00000130203", "ENSG00000142798", "ENSG00000178209", "ENSG00000172889", "ENSG00000147813", "ENSG00000178814", "ENSG00000159069", "ENSG00000182871", "ENSG00000188157", "ENSG00000176978", "ENSG00000167701", "ENSG00000099795", "ENSG00000196924", "ENSG00000176919", "ENSG00000138080", "ENSG00000179403", "ENSG00000164880", "ENSG00000178685", "ENSG00000008710", "ENSG00000182327", "ENSG00000118137", "ENSG00000107331", "ENSG00000166183", "ENSG00000099624", "ENSG00000162337")
# for (gene in black_list) {
#   countstable <- countstable[(countstable[,1] != gene), ]
# }

# Save the gene list, to be used later as row names:
generows <- unlist(countstable[,1], use.names=FALSE)
generows <- generows[-1]
droplevels(generows)

# This column contains the gene names, we don't need it anymore:
countstable[,1] <- NULL

# To sort the table by sample name:
countstable.sorted <- countstable[,order(countstable[1,])]

# Save the samples list, to be used later as column names:
genecols <- unlist(countstable.sorted[1,], use.names=FALSE)
droplevels(genecols)

# To convert the countstable from text to numbers, any string should be removed (names on the first row):
countstable.sorted <- countstable.sorted[-1,]

# Convert the table from characters to numeric:
countstable.num <- as.data.frame(apply(countstable.sorted, c(1,2), as.integer))

# Then, add the row and column names:
rownames(countstable.num) <- generows
colnames(countstable.num) <- genecols

# This line is to filter the countstable, only necessary if you're working with a subset of the sampledata:
countstable.num <- countstable.num[sampledata$id]

# To remove empty rows:
countstable.num <- countstable.num[ rowSums(countstable.num) > 0, ]
```

The following block is only necessary if you wanna apply an independent prior filtering before the differential expression analysis
```{r, echo=FALSE}
##########################################
# Independent prior filtering (optional) #
##########################################

# To remove any gene with 0 reads in at least one sample (i.e. considering only transcripts present in all the samples)
# countstable.num <- cbind(countstable.num, 1)
# for (i in 1:(ncol(countstable.num) - 1)) {
#   countstable.num$`1` <- countstable.num$`1` * countstable.num[,i]
# }
# countstable.num <- countstable.num[countstable.num$`1` != 0,]
# countstable.num$`1` <- NULL

# To work using only transcripts with more than 5 reads in all the samples
# countstable.num <- countstable.num[apply(countstable.num, 1, function(row) (all(row > 5))),]
```

## Data Analysis

### Create the DEseq2 object
The model should be especified in this step

```{r, echo=FALSE}
#########
# DESeq #
#########

# create a DESeqDataSet from the tables above
ds.deseq <- DESeqDataSetFromMatrix(
  countData = countstable.num,
  colData = sampledata,
  design = ~ library_prep + patient + sample_type)
  #design = ~ library_prep + normal)
```

### Comparison
```{r, echo=FALSE}
# Run DESeq2
ds.deseq <- DESeq(ds.deseq)



#This step is to indicate which is the base level (normal)

# This will show the possible comparisons, according to the model set before
resultsNames(ds.deseq)

# p value
cutoff_alpha <- 0.05

# Get the result of a certain comparison (see results help for info on this)
#res <- results(ds.deseq, contrast=c("sample_type", "primary", "normal"),
#res <- results(ds.deseq, contrast=c("normal", "N", "Y"),
res <- results(ds.deseq, contrast=c("sample_type", "metastasis", "primary"),
               altHypothesis="greaterAbs",
               alpha = cutoff_alpha,
               pAdjustMethod = "BH") # Benjamini Hochberg, or FDR

summary(res)
```

### Annotation and output tables
This block is to add the gene names and description.
```{r, echo=FALSE}
#####################
# Adding gene names #
#####################

res$ensembl <- rownames(res)
ensembl = useMart("ensembl", dataset = "hsapiens_gene_ensembl")
# More annotations could also be added from biomaRt:
# listAttributes(mart = ensembl)

genemap <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "description"),
                 filters = "ensembl_gene_id",
                 values = res$ensembl,
                 mart = ensembl)

idx <- match(res$ensembl, genemap$ensembl_gene_id)
res$symbol <- genemap$hgnc_symbol[idx]
res$description <- genemap$description[idx]

```

This is to visualize a summary of the results and make some output tables in .csv format
```{r, echo=FALSE}
#########################################
# Visualizing and filtering the results #
#########################################

summary(res)
#write.csv(as.data.frame(res), file = "results_Tumor_vs_Normal_full.csv")
write.csv(as.data.frame(res), file = "results_Mets_vs_Prims_full.csv")
#write.csv(as.data.frame(res), file = "results_Prims_vs_Normal_full.csv")
head(res, n = 10)
table(res$padj < cutoff_alpha)
table(abs(res$log2FoldChange) >= 1)
table((abs(res$log2FoldChange) >= 1) & (res$padj < cutoff_alpha))

res_sig <- subset(res, ( (padj < cutoff_alpha) & !is.na(padj) ))
summary(res_sig)
head(res_sig, n = 10)
#write.csv(as.data.frame(res), file = "results_Mets_vs_Prims_sig.csv")

res_fold.over <- subset(res_sig, log2FoldChange >= 1)
res_fold.under <- subset(res_sig, log2FoldChange <= -1)

res_fold.over.sorted <- res_fold.over[order(-res_fold.over$log2FoldChange, na.last=TRUE), ]
res_fold.under.sorted <- res_fold.under[order(res_fold.under$log2FoldChange, na.last=TRUE), ]

head(res_fold.over.sorted, n = 10)
head(res_fold.under.sorted, n = 10)

write.csv(as.data.frame(res_fold.over.sorted), file = "results_Mets_vs_Prims_over_(pval_filtered)_(fold_sorted).csv")
write.csv(as.data.frame(res_fold.under.sorted), file = "results_Mets_vs_Prims_under_(pval_filtered)_(fold_sorted).csv")
```

## Variance stabilized transformation
```{r, echo=FALSE}
vsd <- varianceStabilizingTransformation(ds.deseq)
head(assay(vsd))
```

## rLog transformation
```{r, echo=FALSE}
rld <- rlog(dds)
head(assay(rld))
```
